Supplementary Information

### 2021-08-06 doc/manuscript/response_to_reviews.Rmd

 L222-224:
Dealing with tree and forest mortality is difficult. However, climate change is predicted to affect mortality in many forests. In turn, mortality can impact the growth of the remaining trees depending on the stature and location of the dead tree. This point should be addressed- the article presents the relationship between growth and climate only for the surviving trees. How does excluding growth data from the analysis potentially bias the results?

*increased mortality--> decreased competition --> increased growth. This should be mentioned in the discussion.*

P19-20 (especially L301-303):
I think it could be made more clear that detrending was used only in the process of selecting strongly related climate variables, and not in the GLS models. Additionally, a comment on how the variability in the application of detrending method used here (thin plate splines) would impact climate variable selection would be warranted. There is a persistent question in dendrochronology about the amount of climate signal gets removed through the process of detrending.

**In the GLS models, we use a temporal autocorrelation structure on raw data. We have revised the GLS model description to make this more clear: "We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*). We will refer to this model as a generalized least squares (GLS) model (Fig. 1).Within the GLS models, our response variables were raw, log-transformed growth estimates (as opposed to residuals): log[RW], log[BAI], or log[$\Delta$ AGB].**

**Dentrending method is often a highly contentious topic within the dendro community, especially when reconstructing climate variables over long time scales . In our study, our candidate variables had large interannual variability and the relatively short series length does not extend to the point where detrending method will have a significant influence on the identified climate signal [@bunn_using_2004].  We have compared our approach with more traditional detrending techniques and observed that there were not statistically significant differences among the methodologies (Appendix S4). To clarify this in the manuscript, we have added the following sentence (under *Step 1:...* header): "Our application of the thin-plate regression splines acts similar to more traditional a priori detrending methods using a two-thirds spline commonly used in dendrochronology studies and results in similar predictor variable selection [Appendix 4, @cook_calculating_1997; @rollinson_climate_2021]."**

@@

L94: enormous is a subjective word, change

**We have changed this word to "large".**

L96-100: sentence awkward (not sure how the “the or to predict...” follows from first part of sentence)

**We have reworded as follows: "Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees -- including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @rollinson_climate_2021] -- or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020]."**

@@

L269: Defined ‘most important’

**We have reworded as follows: "We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most influential climate variable (*i.e.*, that most strongly correlated with annual growth)..."**

@@

L280: linear and quadratic terms?

**We have changed this sentence to read, "...specifying linear and quadratic terms to allow for potential nonlinearities in the climate response."**

L308: So models with variables removed may include one climate variable and DBH?

**In theory, yes. In practice, no variables required removing. To make this more clear, we have reworded as follows: "Prior to running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014]. Our analysis code was programmed to remove any variable with a variance inflation factor > 3, but none required removal."**

@@

L311: Clarify “we ran every combination”. This is only for selected climate variable right?

**We have reworded as follows: "For each species independently, we ran GLS models including every possible combination of independent variables (*i.e.*, candidate climate drivers, DBH, and year), including both first- and second-order terms for each."**

@@

L627: What are you reconstructing? You mean back-calculating DBH? Probably a terminology thing.

**"Reconstructing" refers primarily to  back-calculating, although in some cases DBH is projected slightly forward (e.g., at SCBI, where cores were taken a couple years after a census was conducted). For this reason, we prefer to stick with the "reconstructed" term. To make sure this is clear, we now refer to the term "back-calculated" in the methods section when we introduce the concept: " For each year in the tree-ring records, we reconstructed (*i.e.*, back-calculated) DBH, as detailed in Appendix S2."**

L678, 683: Weird formatting issue in ref. Maybe journal software?

**I do not see any problem (on a Mac), including in the draft proof. I have opened a GitHub issue (https://github.com/EcoClimLab/ForestGEO-tree-rings/issues/142) to remind myself to check this at the proof stage.**

### 2021-08-06 doc/manuscript/ForestGEO_treerings.Rmd
worked on reviewer responses

Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989; @zuidema_tropical_2013]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002].
Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
Tree rings also provide the long-term perspective necessary for understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2021].
This information is critical to predicting forest responses to anthropogenic changes -- particularly climate change -- and thereby reducing the large uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @rollinson_climate_2021] or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest carbon sequestration in the current era of rapid environmental change.

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
These types of detrending methods are used to maximize the climate signal in tree-ring paleoclimate reconstructions and thus have been shown to retain the climate signal while removing the influence of other non-climatic drivers [@fritts_tree_1976; @cook_time_1985].  
We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying linear and quadratic terms to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination with the lowest AIC as a candidate climate variable for the multivariate models.

@@

We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*). 
We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).

Prior to running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
Within the GLS models, our response variables were log[RW], log[BAI], or log[$\Delta$ AGB].

For each species independently, we ran every combination of the candidate climate drivers and DBH, including both first- and second-order terms for each.
For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that exponential functions would be captured by a linear fit to the log-transformed growth variables, while u-shaped fits are not expected biologically.

>>>>

Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989; @zuidema_tropical_2013]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002].
Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
Tree rings also provide the long-term perspective necessary for understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2021].
This information is critical to predicting forest responses to anthropogenic changes -- particularly climate change -- and thereby reducing the large uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees -- including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @rollinson_climate_2021] -- or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest carbon sequestration in the current era of rapid environmental change.

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most influential climate variable (*i.e.*, that most strongly correlated with annual growth) and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
Our application of the thin-plate regression splines acts similar to more traditional a priori detrending methods using a two-thirds spline commonly used in dendrochronology studies and results in similar predictor variable selection [Appendix `r traditional_comparison`, @cook_calculating_1997; @rollinson_climate_2021].  
We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying linear and quadratic terms to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination with the lowest AIC as a candidate climate variable for the multivariate models.

@@

We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*). 
We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).
Within the GLS models, our response variables were raw, log-transformed growth estimates (as opposed to residuals): log[RW], log[BAI], or log[$\Delta$ AGB].
Prior to running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014]. Our analysis code was programmed to remove any variable with a variance inflation factor > 3, but none required removal.

For each species independently, we ran GLS models including every possible combination of independent variables (*i.e.*, candidate climate drivers, DBH, and year), including both first- and second-order terms for each.
For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that exponential functions would be captured by a linear fit to the log-transformed growth variables, while u-shaped fits are not expected biologically.
